---
title: 'MetaboDiff tutorial'
author: | 
  | Andreas Mock
  | Experimental Neurosurgery, Heidelberg University Hospital,
  | Division of Applied Bioinformatics, German Cancer Research Center (DFKZ) Heidelberg &
  | Department of Medical Oncology, National Center for Tumor Diseases (NCT) Heidelberg
date: '`r Sys.Date()`'
output: BiocStyle::html_document
fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Cancer Variant Explorer (CVE) tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{utf8}
---

```{r setup, include=FALSE}
library("devtools")
library("gdata")
library("SummarizedExperiment")
install_github("andreasmock/MetaboDiff")
library("MetaboDiff")
library("tidyverse")
library("ggfortify")
library("ComplexHeatmap")
```

# Introduction

Comparative metabolomics comes of age by an increasing list of commercial vendors (i.e. Metabolon &reg;) offering reproducible high-quality metabolomic data for translational researchers outside the mass spectrometry field. This R packages aims to provide a low-level entry to differential metabolomic analysis by starting off with the table of relative metabolite quantifications provided by commercial vendors.

# Installation

The `MetaboDiff` R package can be installed via Github

```{r,eval=FALSE}
library(devtools)
install_github("andreasmock/MetaboDiff")
```

and once installed loaded by

```{r,eval=FALSE, message=FALSE}
library("MetaboDiff")
```

# Reading data and annotation

## Example data

The example data is derived from a study by Priolo and colleagues in which they used the service of the Metabolon&reg; company to compare the tissue metabolome of 40 prostate cancers with 16 normal prostate specimens.[^1] 
The table summarizing the relative metabolic measurements can be loaded from the journal homepage as follows:

```{r, eval=TRUE, message=FALSE}
data = as.matrix(read.xls("http://cancerres.aacrjournals.org/highwire/filestream/290852/field_highwire_adjunct_files/4/131853_1_supp_2658512_nbpsn6.xlsx",sheet=5,na.strings = ""))
```

## Input files

To ensure a seamless analysis, the metabolomic data is stored as an `SummarizedExperiment` object [^2]. Hence, the following input object are needed 

- `assay` - a matrix containing the relative metabolic measurements.
- `colData` - a dataframe containing the sample metadata.
- `rowData` - a dataframe containing the metabolite annotation.


### Creation of assay object containing metabolomic data

```{r, eval=TRUE}
assay = apply(data[3:nrow(data),8:ncol(data)],2,as.numeric)
assay[1:4,1:5]
```

### Creation of colData object containing sample metadata

```{r}
colData = data.frame(sample_id = colnames(t(data[1,8:ncol(data)])),
                     tumor_normal = as.vector(t(data[1,8:ncol(data)])))
head(colData)
```

### Creation of rowData object containing metabolite annotations and ids 

For the rowData object we start off with the annotation already provided by the commercial vendor: 

```{r}
rowData = as.data.frame(data[3:nrow(data),1:7])
colnames(rowData) = data[2,1:7]
colnames(rowData)[7] = "HMDB_ID"
colnames(rowData)
```

### Annotation using Small Molecular Pathway Database (SMPDB)

Alongside the metabolic measurments, Metabolon &reg; provides metabolite annotation including so called super-pathways and sub-pathways. However, theses annotations might very from vendor to vendor. Hence, the `MetaboDiff` package makes use of the Small Molecular Pathway Database (SMPDB) for metabolite annotation[^3]. `MetaboDiff` supports all common metabolic IDs as input for the annotation (HMDB, KEGG and ChEBI). In the example data, both the KEGG and the HMDB identifier are available. As the databases (HMDB, KEGG or ChEBI) cover unique annotations due to different standards in identifying and reporting metabolites [^4], the function `get_SMPDBanno` queries the database using all three ids (if available) and joins all available information. The current SMPDB build used within `MetaboDiff` is version 2.0 and will be updated as new versions are released. Please refer to the development branch of `MetaboDiff` to work with the latest SMPDB annotation.

```{r}
rowData = get_SMPDBanno(rowData,
                        column_kegg_id=6,
                        column_hmdb_id=7,
                        column_chebi_id=NA)
```

## Create SummarizedExperiment object

```{r}
met = SummarizedExperiment(assays=assay,
                           rowData=rowData,
                           colData=colData)
met
```

# Imputation of missing values

```{r}
met_na = is.na(assay(met))*1
row_na = apply(met_na,1,sum)/nrow(met_na)
col_na = apply(met_na,2,sum)/ncol(met_na)

rowanno = rowAnnotation(barplot=row_anno_barplot(row_na[order(row_na)],
                                                 gp=gpar(fill="brown",
                                                         col="brown"),
                                                 axis = TRUE),
                        width=unit(1.5,"cm"))
colanno = columnAnnotation(barplot=anno_barplot(col_na[order(col_na)],
                                                 gp=gpar(fill="brown",
                                                         col="brown")))

ha = Heatmap(matrix = met_na[order(row_na),order(col_na)],
        name="missing",
        col=c("lightgrey","brown"),
        cluster_row = FALSE,
        cluster_columns = FALSE,
        column_title = "samples",
        column_title_side = "bottom",
        show_column_names = FALSE,
        heatmap_legend_param = list(labels=c("no","yes")),
        row_title= "metabolites",
        bottom_annotation = colanno,
        bottom_annotation_height = unit(1.5,"cm"))
ha + rowanno
```


# Data preprocessing and quality control

```{r}
#reshape data for ggplot2
mdata = gather(as.tibble(assay(met)[,1:20]))
ggplot(mdata, mapping=aes(x=key,y=value)) + 
    geom_boxplot() 
```

# Unsupervised analysis

```{r}
autoplot(prcomp(as.tibble(assay(met))))
```


# Compartative analysis

# Visualization of metabolic networks

To compare the structure of metabolic networks across tumor entities, network visualizations should ideally be perceptual uniform. However, networks are usually visualized using force-directed layouts (e.g. Fruchterman Reingold[^3]) generating pleasant but hardly interpretable networks earning the nickname ’hairballs’.

## Hive plots

To address this issue, Kryzwinski et al. have developed an algorithm in which nodes are placed on radially oriented linear axes according to a well-defined coordinate system[^4]. In resemblance to the top of a beehive, the authors termed these visualizations hive plots and show that they can be adjusted to suit the purpose, are easily explained and understood and most importantly can generate useful and quantitatively inter- pretable results.

# Differentially abundant subpathways

graph clustering - hotnet2 

# Session information

```{r}
sessionInfo()
```

# References
[^1]: Priolo, C., Pyne, S., Rose, J., Regan, E. R., Zadra, G., Photopoulos, C., et al. (2014). AKT1 and MYC Induce Distinctive Metabolic Fingerprints in Human Prostate Cancer. Cancer Research, 74(24), 7198–7204. http://doi.org/10.1158/0008-5472.CAN-14-1490

[^2]: M. Morgan, V. Obenchain, J. Hester, and P. Herv, “Summarizedexperiment: Summarized- experiment container,” R package version, vol. 1.2.3, 2016.

[^3]: Jewison T, Su Y, Disfany FM, et al. SMPDB 2.0: Big Improvements to the Small Molecule Pathway Database Nucleic Acids Res. 2014 Jan;42(Database issue):D478-84.

[^4]: Thiele, I., Swainston, N., Fleming, R. M. T., Hoppe, A., Sahoo, S., Aurich, M. K., et al. (2013). A community-driven global reconstruction of human metabolism. Nature Biotechnology, 31(5), 419–425. http://doi.org/10.1038/nbt.2488

[^5]: T. M. J. Fruchterman and E. M. Reingold, “Graph drawing by force-directed placement,” Software: Practice and Experience, vol. 21, no. 11, pp. 1129–1164, 1991.

[^6]: M. Krzywinski, I. Birol, S. J. M. Jones, and M. A. Marra, “Hive plots–rational approach to visualizing networks.,” Briefings in Bioinformatics, vol. 13, pp. 627–644, Sept. 2012.


