---
title: 'MetaboDiff tutorial'
author: | 
  | Andreas Mock
  | Experimental Neurosurgery, Heidelberg University Hospital,
  | Division of Applied Bioinformatics, German Cancer Research Center (DFKZ) Heidelberg &
  | Department of Medical Oncology, National Center for Tumor Diseases (NCT) Heidelberg
date: '`r Sys.Date()`'
output:
  github_document:
    toc: true
    toc_depth: 2
fig_caption: yes
---

```{r setup, include=FALSE}
library("devtools")
library("gdata")
library("MultiAssayExperiment")
library("SummarizedExperiment")
library("MetaboDiff")
library("tidyverse")
library("ggfortify")
library("ComplexHeatmap")
library("impute")
library("vsn")
library("HiveR")
library("RColorBrewer")
library("plyr")
library("cowplot")
library("cluster")
library("factoextra")
install_github("andreasmock/MetaboDiff")
```

# Introduction

Comparative non-targeted metabolomics comes of age through an increasing number of commercial vendors offering reproducible high-quality metabolomic data for translational researchers outside the mass spectrometry field. This R packages aims to provide a low-level entry to differential metabolomic analysis by starting off with the table of relative metabolite quantifications provided by commercial vendors or core facilities. 

# Installation

The `MetaboDiff` R package can be installed via Github

```{r,eval=FALSE}
library(devtools)
install_github("andreasmock/MetaboDiff")
```

and once installed loaded by

```{r,eval=FALSE, message=FALSE}
library("MetaboDiff")
```

# Part I: Data processing

## Reading data and annotation

### Example data

The example data is derived from a study by Priolo and colleagues in which they used the service of Metabolon&reg; to compare the tissue metabolome of 40 prostate cancers with 16 normal prostate specimens.[^1] 
The table summarizing the relative metabolic measurements can be loaded from the journal homepage as follows:

```{r, eval=TRUE, message=FALSE}
data = as.matrix(read.xls("http://cancerres.aacrjournals.org/highwire/filestream/290852/field_highwire_adjunct_files/4/131853_1_supp_2658512_nbpsn6.xlsx",sheet=5,na.strings = ""))
```

### Input files

The metabolomic data within `MetaboDiff` is stored as a `MultiAssayExperiment`class [^2]. This framework enables the coordinated representation of multiple experiments on partially overlapping samples with associated metadata and integrated subsetting across experiments. In the context of metabolomic data analysis, multiple assays are needed to store raw data and imputed data which usually contain different number of metabolites due to missing values (see section on data imputation for more details). 

The core components of the `MultiAssayExperiment` class are:

* `ExperimentList` - a slot of class ExperimentList containing data for each experimental assay. Within the ExperimentList slot, the metabolomic data is stored as a 
SummarizedExperiment object consisting of:
    + `assay` - a matrix containing the relative metabolic measurements. 
    + `rowData` - a dataframe containing the metabolite annotation..
* `colData` - a slot of class data frame describing the sample metadata available across all experiments
* `sampleMap` - a slot of class data frame relating clinical data to experimental assay

#### Creation of assay object containing metabolomic data

```{r, eval=TRUE}
assay = apply(data[3:nrow(data),8:ncol(data)],2,as.numeric)
colnames(assay) = paste0(rep("sample",ncol(assay)),1:ncol(assay))
rownames(assay) = paste0(rep("met",nrow(assay)),1:nrow(assay))
assay[1:4,1:5]
```

#### Creation of colData object containing sample metadata

```{r}
colData = data.frame(id = colnames(t(data[1,8:ncol(data)])),
                     tumor_normal = as.vector(t(data[1,8:ncol(data)])),
                   row.names=paste0(rep("pat",ncol(assay)),1:ncol(assay)))
head(colData)
```

#### Creation of rowData object containing metabolite annotations and ids 

For the rowData object we start off with the annotation already provided by the commercial vendor: 

```{r}
rowData = as.data.frame(data[3:nrow(data),1:7])
colnames(rowData) = data[2,1:7]
colnames(rowData)[7] = "HMDB_ID"
rownames(rowData) = paste0(rep("met",nrow(assay)),1:nrow(assay))
colnames(rowData)
```

#### Annotation using Small Molecular Pathway Database (SMPDB)

Alongside the metabolic measurements, Metabolon &reg; provides metabolite annotation including so called super-pathways and sub-pathways. However, theses annotations might very from vendor to vendor. Hence, the `MetaboDiff` package makes use of the Small Molecular Pathway Database (SMPDB) for metabolite annotation[^3]. `MetaboDiff` supports all common metabolic IDs as input for the annotation (HMDB, KEGG and ChEBI). In the example data, both the KEGG and the HMDB identifier are available. As the databases (HMDB, KEGG or ChEBI) cover unique annotations due to different standards in identifying and reporting metabolites [^4], the function `get_SMPDBanno` queries the database using all three ids (if available) and joins all available information. The current SMPDB build used within `MetaboDiff` is version 2.0 and will be updated as new versions are released. Please refer to the development branch of `MetaboDiff` to work with the latest SMPDB annotation.

```{r}
rowData = get_SMPDBanno(rowData,
                        column_kegg_id=6,
                        column_hmdb_id=7,
                        column_chebi_id=NA)
```

### Creation of SummarizedExperiment object and ExperimentList

```{r}
se = SummarizedExperiment(assays=assay,
                           rowData=rowData)
experiment_list = list(raw=se)
experiment_list
```

### Creation of sampleMap 

```{r}
sampleMap = data.frame(primary=rownames(colData),
                 colname=colnames(se))
sampleMap_list = listToMap(list(raw=sampleMap))
```

### Construction of MultiAssayExperiment

```{r, message=FALSE}
met = MultiAssayExperiment(experiments = experiment_list,
                           colData = colData,
                           sampleMap = sampleMap_list)
met
```

## Imputation of missing values

In contrast to other high-throughput technologies, missing values are common in quantitative metabolomic datasets. 

The function `na_heatmap` visualizes the missing metabolite measurements across the samples. A vector of sample labels (e.g. tumor vs. normal) and the color of the sample levels needs to be specified.

```{r}
na_heatmap(met,
           sample_labels=colData(met)$tumor_normal,
           label_colors=c("darkseagreen","dodgerblue"))
```

The example data supports the need for data imputation. It could be shown that k-nearest neighbor imputation minimizes the effects on the normality and variance of the data as long as the number of missing data does not exceed 40% [^5]. 

The function 'knn_impute' adds the slot "impute" to the MultiAssayExperiment object that contains the imputed relative metabolite measurements for all metabolites with raw measurements in more than 60% of cases. We recommend a cutoff of 40% (i.e. 0.4). However the cutoff might be changed according to the discretion of the user.

```{r}
met = knn_impute(met,cutoff=0.4)
met
```

As apparent form the summary description of the object 'met' the imputed data matrix contains only 238 of the 307 original metabolites according the cutoff of 40% missing values.

## Removal of outliers

Before we normalize the data, we want to exclude outliers in the study set. To this end, we provide the function `outlier_heatmap`.
The sample annotation shows the number of missing metabolites per sample as the proxy of the impact of imputation on clustering. 
To showcase outliers, the hierarchical clustering tree is cut in 2 clusters.

```{r}
outlier_heatmap(met,
                sample_labels=colData(met)$tumor_normal,
                label_colors=c("darkseagreen","dodgerblue"))
```

The imputed data of the example study set displays a cluster of 5 samples (cluster 1) with an in average lower relative metabolite abundance. Due to the lack of batch information, this cannot be investigted further at this time. To demonstrate, how a cluster can be removed, we apply the function `remove_cluster` to remove cluster 1:

```{r}
met = remove_cluster(met,cluster=1)
met
```

As displayed in the summary of the `met` object, the 5 samples of cluster 1 were successfully removed from the slots "raw" and "imputed".

## Normalization

Variance stabilizing normalization (vsn) is used ensure that the variance remains nearly constant over the measured spectrum (Huber et al., 2002).

```{r}
met = normalize_met(met)
```

## Summary of processed object

```{r}
met
```

At this point the data processing is completed. The `MultiAssayExperiment` object contains now 4 slots:

- raw - raw relative metabolic measurements as provided by company or core facility
- imputed - imputed relative metabolic measurements (k-nearest neighbor imputation)
- norm - normalized relative metabolic measurements (vsn)
- norm_imputed - normalized imputed relative metabolic measurements (vsn)

## Quality control of normalization

```{r,fig.width=10,fig.height=8, message=FALSE}
quality_plot(met)
```

# Part II: Data analysis

## Unsupervised analysis

### PCA plot

```{r}
autoplot(prcomp(t(assays(met)[["norm_imputed"]]),center = TRUE,scale = FALSE),
         data=as.data.frame(colData(met)),
         colour="tumor_normal")
```


### Correlation heatmap

### Consensus clustering

## Comparative analysis



# Session information

```{r}
sessionInfo()
```

# References
[^1]: Priolo, C., Pyne, S., Rose, J., Regan, E. R., Zadra, G., Photopoulos, C., et al. (2014). AKT1 and MYC Induce Distinctive Metabolic Fingerprints in Human Prostate Cancer. Cancer Research, 74(24), 7198–7204. http://doi.org/10.1158/0008-5472.CAN-14-1490

[^2]: Sig M (2017). MultiAssayExperiment: Software for the integration of multi-omics experiments in Bioconductor. R package version 1.2.1

[^3]: Jewison T, Su Y, Disfany FM, et al. SMPDB 2.0: Big Improvements to the Small Molecule Pathway Database Nucleic Acids Res. 2014 Jan;42(Database issue):D478-84.

[^4]: Thiele, I., Swainston, N., Fleming, R. M. T., Hoppe, A., Sahoo, S., Aurich, M. K., et al. (2013). A community-driven global reconstruction of human metabolism. Nature Biotechnology, 31(5), 419–425. http://doi.org/10.1038/nbt.2488

[^5]: Armitage, E. G., Godzien, J., Alonso-Herranz, V., L pez-Gonz lvez, N., & Barbas, C. (2015). Missing value imputation strategies for metabolomics data. Electrophoresis, 36(24), 3050–3060. http://doi.org/10.1002/elps.201500352

[^6]: T. M. J. Fruchterman and E. M. Reingold, “Graph drawing by force-directed placement,” Software: Practice and Experience, vol. 21, no. 11, pp. 1129–1164, 1991.

[^7]: M. Krzywinski, I. Birol, S. J. M. Jones, and M. A. Marra, “Hive plots–rational approach to visualizing networks.,” Briefings in Bioinformatics, vol. 13, pp. 627–644, Sept. 2012.


